Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. In contrast,
conditioning involves restricting the focus to the subset of the
population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).\\

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \:\bot\:e_{X} \\
  e_{T} & \:\bot\:e_{Y} \\
  e_{X} & \:\bot\:e_{Y}
\end{aligned}
\]

@@

Beyond their graphical capabilities, SCMs and DAGs can encode the
probabilistic information embedded within a causal structure. They
achieve this encoding by relying on three fundamental assumptions: the
local Markov, the minimality, the causal edges assumptions. The
\emph{local Markov assumption} encodes probabilistic independencies
between variables by declaring that nodes in a graph are independent of
all its non-descendants, given its parents \citep[pp.~20]{Neal_2020}.
Meanwhile, the \emph{minimality assumption} encodes probabilistic
dependencies among variables by stating that every pair of adjacent
nodes exhibits a dependency \citep[pp.~21]{Neal_2020}. Finally, the
\emph{causal edges assumption} encodes causal relationships between
variables by declaring that each parent node acts as a direct cause of
its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
illustrates how these assumptions influence the statistical and causal
interpretations of graphs.

@@

A notable implication of the assumptions underlying the probabilistic
encoding is that any conceptual model described by an SCM and DAG can
represent the joint distribution of variables more efficiently
\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
product of conditional probability distributions (CPDs)
\(P(child \:|\:parents)\). This property is formally known as the
\emph{Bayesian Network factorization} (BNF,
Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this expression, \(pa(X_{i})\) denotes the set of variables that are
the parents of \(X_{i}\).

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-C} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
supports the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\), which defines probabilistic
independence \citep[pp.~24]{Neal_2020}. Conversely, when \(X_{3}\) is
unconditionally dependent on \(X_{1}\) (\(X_{1} \:\not\bot\:X_{3}\)), as
depicted in the SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}, the
BNF express their joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\). Notably, these
descriptions demonstrate the clear correspondence between the structural
equations illustrated in the figures of Section~\ref{sec-appendix-C} and
the CPDs.

Beyond the insights gained from two-node structures, researchers can
uncover more nuanced patterns of (in)dependencies from chains, forks,
and colliders. For instance, researchers applying the BNF to the chain
structure depicted in the SCM \ref{fig-scm_bb3} and DAG
\ref{fig-mdag_bb3} can represent the joint distribution for the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). From this
expression, researchers can infer additional (in)dependencies assumed in
any data generated by this structure. Specifically, the structure
implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). In the
fork structure shown in the SCM \ref{fig-scm_bb4} and DAG
\ref{fig-mdag_bb4}, researchers can express the joint distribution of
the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
conditionally independent when controlling for \(X_{2}\)
\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} can express the joint distribution of the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
dependent when controlling for \(X_{2}\)
\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
mathematical proofs for these conclusions.

@@

\begin{equation}\phantomsection\label{eq-trunc_factor}{
P(X_{1}, X_{2}, \dots, X_{P} \:|\:do(S)) =
\begin{cases}
  \prod P(X_{i} \:|\:pa(X_{i}) ) & \text{if} \: p \not\in S \\
  1 \quad & \text{otherwise}
\end{cases}
}\end{equation}

@@

@@

\begin{minipage}{0.33\linewidth}

@@

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}

>>>>

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. In contrast,
\emph{conditioning} involves restricting the focus to the subset of the
population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).\\

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{T}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \:\bot\:e_{X} \\
  e_{T} & \:\bot\:e_{Y} \\
  e_{X} & \:\bot\:e_{Y}
\end{aligned}
\]

@@

Beyond their graphical capabilities, SCMs and DAGs can encode the
probabilistic information embedded within a causal structure. They
achieve this encoding by relying on three fundamental assumptions: the
local Markov, the minimality, the causal edges assumption. The
\emph{local Markov assumption} encodes probabilistic independencies
between variables by declaring that nodes in a graph are independent of
all its non-descendants, given its parents \citep[pp.~20]{Neal_2020}.
Meanwhile, the \emph{minimality assumption} encodes probabilistic
dependencies among variables by stating that every pair of adjacent
nodes exhibits a dependency \citep[pp.~21]{Neal_2020}. Finally, the
\emph{causal edges assumption} encodes causal relationships between
variables by declaring that each parent node acts as a direct cause of
its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
illustrates how these assumptions influence the statistical and causal
interpretations of graphs.

@@

A notable implication of the assumptions underlying the probabilistic
encoding is that any conceptual model described by an SCM and DAG can
represent the joint distribution of variables more efficiently
\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
product of conditional probability distributions (CPDs) of the type
\(P(child \:|\:parents)\). This property is formally known as the
\emph{Bayesian Network factorization} (BNF,
Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this expression, \(pa(X_{i})\) denotes the set of variables that are
the parents of \(X_{i}\).

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-C} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
supports the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent or d-separated
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-C} and
the CPDs.

Beyond the insights gained from two-node structures, researchers can
uncover more nuanced patterns of (in)dependencies from chains, forks,
and colliders. These (in)dependencies apply to any data generated by
these structures. For instance, applying the BNF to the chain structure
depicted in the SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow
researchers to represent the joint distribution for the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
conditionally independent when controlling for \(X_{2}\)
\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} can express the joint distribution of the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
dependent when controlling for \(X_{2}\)
\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
mathematical proofs for these conclusions.

Researchers can revisit the scenario in Section~\ref{sec-appendix-B}
using these insights. By applying the BNF to the SCM
\ref{fig-scm_example1} and DAG \ref{fig-mdag_example1a} structures, they
can represent the joint probability of the observed variables as
\(P(Y, T, X) = P(Y \:|\:T, X) P(T \:|\:X) P(X)\). Assuming the
\emph{direct effect} of \(T\) on \(Y\) (\(T \rightarrow Y\), black path)
is absent, researchers can infer from the joint probability that \(Y\)
is not unconditionally independent of \(T\) \((Y \:\bot\:T)\). This
dependency arises because a non-causal path exists between \(T\) and
\(Y\) through \(X\) (\(T \leftarrow X \rightarrow Y\), gray path),
indicating that \(X\) is a confounder. However, researchers can observe
that conditioning on \(X\), depicted as a square in DAG
\ref{fig-mdag_example1a}, results in \(Y \:\bot\:T \:|\:X\), eliminating
the confounding effect and allowing the identification of the direct
effect \(T \rightarrow Y\) if it exists.

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{T}(X,e_{T}) \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \:\bot\:e_{X} \\
  e_{T} & \:\bot\:e_{Y} \\
  e_{X} & \:\bot\:e_{Y}
\end{aligned}
\]

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}

@@

In addition to the assumptions required to encode probabilistic
information, SCMs and DAGs rely on the modularity assumption to
represent interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
Section~\ref{sec-appendix-B}). Defining these interventions enables
researchers to identify the ACE from associational quantities through
statistical conditioning, effectively disentangling causation from mere
association.

The modularity assumption underpins the concepts of Truncated
Factorization and manipulated graphs, which are essential for
representing interventions within SCMs and DAGs. The \emph{modularity
assumption} posits that intervening on a node modifies only the causal
mechanism of that node, leaving others unchanged
\citep[pp.~34]{Neal_2020}. Building on this principle, the
\emph{Truncated Factorization} (TF) simulates physical interventions by
removing specific functions from the conceptual model and substituting
them with constants, preserving the rest of the structure unchanged
{[}Pearl\_2010{]}. The probabilistic implications of this factorization
are formalized in Equation~\ref{eq-trunc_factor}, where \(S\) denotes
the subset of variables \(X_{p}\) directly influenced by the
intervention. Complementing the TF, \emph{manipulated graphs} are DAGs
where specific edges have been removed to reflect the effects of
interventions \citep[pp.~34]{Neal_2020}.

\begin{equation}\phantomsection\label{eq-trunc_factor}{
P(X_{1}, X_{2}, \dots, X_{P} \:|\:do(S)) =
\begin{cases}
  \prod P(X_{p} \:|\:pa(X_{p}) ) & \text{if} \: p \not\in S \\
  1 \quad & \text{otherwise}
\end{cases}
}\end{equation}

Using the TF, researchers can define the \emph{backdoor adjustment}, a
method for identifying causal effects. This adjustment specifies that if
a variable \(X_{p} \in S\) serves as a \emph{sufficient adjustment set}
for the effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified
using Equation~\ref{eq-backdoor} (see Section~\ref{sec-appendix-B}).

@@

Ultimately, backdoor adjustment enables researchers to express the ACE
as:

@@

@@

@@

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}
